Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2017-12-09 paper.tex
CL

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for the proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666
%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The ligand is modeled using the set $\L$ of its conformations.
Candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
Each candidate conformation is ranked according its energy and only low-energy conformations are used in $\L$.
%The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
%with energy lower than a user-defined threshold are used.
%The flexibility of the ligand is modeled using the set $\L$ of its conformations.
%We assume that the library of all possible conformations in a considered resolution is available
%(e.g., ~\cite{dunbrack}) 
%and we later show how to select a subset of conformations suitable for the motion planning.
To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
A random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
If the sample $q$ is collision-free, it is added to $\QI$ as a new starting configuration.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
The distance $\RI$ influences how far are the initial configurations from the start of the tunnel.
By placing them too far, the planner would need more iterations to enter the tunnel, which would increase the runtime.
Contrary, too low $\RI$ may disable to find any collision-free configurations, especially for large ligands.
We suggest to set $\RI$ to the radius of the first sphere in the tunnel and increase it only if no collision-free configuration
can be found around.
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

@@

Movements of the ligands inside a tunnel are blocked by many atoms forming the tunnel.
It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
should be allowed, which is controlled by the parameter $\dt$.
This parameter should be set considering existence of cavities or other tunnels around the tunnel being investigated.
In the case of alone tunnels (which is very common), we recommend to use double of the tunnel average width.

@@

The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
$\da(q_1,q_2)$.
This metric is sensitive to the shape of the ligand (that depends on its conformation) as it computes the distance using the two nearest atoms of the two configurations.
It allows us to expand the tree by a configuration whose nearest atom approaches its nearest atom in $\qrand$, which supports
retraction of the ligands towards $\qrand$.

@@

The pool of available conformations $\L$ consists all conformations (upto some resolution) with energy lower than a user specified threshold, which can still contain too many conformations (thousands or even more).
We select a subset by computing how the atoms of the ligand are spread around its geometric mean.
The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
distances then characterizes the spread.
Low deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
To select $n$ conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.

@@

The proposed method was used to compute ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (PDB ID 4E46) which consists of 4650 atoms.
The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
MD of the protein was computed using AMBER~12 (details about the used MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
The tunnel traversability was analyzed in 100 consecutive frames of MD simulation in part where the protein was stabilized.

@@

The performance of the methods is evaluated using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
This distance is measured as 3D distance between geometric center of the ligand and the position of the active site.
%~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}

@@

The aim of this experiment was to verify the procedure to select feasible ligand conformations from an available pool of conformations.
Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
The pool of conformations $\L$ was created by rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
Two strategies to select a subset from the $\L$ conformations were tested: 
the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
to the selection of packed conformations.
The strategies were tested using the ligand DCP (Dichloropropanol), for which $1300$ conformations were created, and for the ligand
m040 (1,5-Dichloropentane) with $7560$ conformations.

>>>>

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for the proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666
%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The ligand is modeled using the set $\L$ of its conformations.
The candidate conformations are prepared by discrete sampling of degrees of freedom of the ligand.
Each candidate conformation is ranked according to its energy and only low-energy conformations are used in $\L$.
%The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
%with energy lower than a user-defined threshold are used.
%The flexibility of the ligand is modeled using the set $\L$ of its conformations.
%We assume that the library of all possible conformations in a considered resolution is available
%(e.g., ~\cite{dunbrack}) 
%and we later show how to select a subset of conformations suitable for the motion planning.
To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
A random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
If the sample $q$ is collision-free, it is added to $\QI$ as a new starting configuration.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
The distance $\RI$ influences how far the initial configurations are from the start of the tunnel.
By placing them too far, the planner would need more iterations to enter the tunnel, which would increase the runtime.
Contrary, too low $\RI$ may disable to find any collision-free configurations, especially for large ligands.
We suggest to set $\RI$ to the radius of the first sphere in the tunnel and increase it only if no collision-free configuration
can be found around.
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

@@

The movements of the ligands inside a tunnel are blocked by many atoms forming the tunnel.
It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
should be allowed, which is controlled by the $\dt$ parameter.
This parameter should be set considering the existence of cavities or other tunnels around the tunnel being investigated.
In the case of alone tunnels (which is very common), we recommend to use double value of the tunnel average width.

@@

The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
$\da(q_1,q_2)$.
This metric is sensitive to the shape of the ligand (that depends on its conformation) as it computes the distance using the two nearest atoms of the two configurations.
It allows us to expand the tree by a configuration whose nearest atom approaches its nearest atom in $\qrand$, which supports
the retraction of the ligands towards $\qrand$.

@@

The pool of available conformations $\L$ consists all conformations (up to some resolution) with energy lower than a user specified threshold, which can still contain too many conformations (thousands or even more).
We select a subset by computing how the atoms of the ligand are spread around its geometric mean.
The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
distances then characterizes the spread.
Low deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
To select $n$ conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.

@@

The proposed method was used to compute the ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (PDB ID 4E46) which consists of 4650 atoms.
The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
MD of the protein was computed using AMBER~12 (details about the used MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
The tunnel traversability was analyzed in 100 consecutive frames of MD simulation in part where the protein was stabilized.

@@

The performance of the methods is evaluated using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
This distance is measured as 3D distance between the geometric center of the ligand and the position of the active site.
%~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}

@@

The aim of this experiment was to verify the procedure to select feasible ligand conformations from an available pool of conformations.
Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
Two strategies to select a subset from the $\L$ conformations were tested: 
the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
to the selection of packed conformations.
The strategies were tested using the ligand DCP (Dichloropropanol), for which $1300$ conformations were created, and for the ligand
m040 (1,5-Dichloropentane) with $7560$ conformations.
